---
title: "Untitled"
format: html
---
```{r, message=FALSE}
#| output: False
#| echo: true
#| eval: false


# name: Will Sovine, Stephen Stull, Travis Darby
# date: 03/25/23
# assignment name: Final Project Code
# goal of program: Conduct EDA on data for the purpose creating decision tree, 
#                  random forest, knn, and logistic regression models to predict
#                  whether or not cells are Maligant or Benign. 
# data files used: FNA_cancer.csv
```

# Data/Library Import
```{r, message=FALSE}

# Import packages
library(tidyverse)
library(tidymodels)
library(rpart)
library(partykit)
library(randomForest)
library(class)
library(caret)
library(MASS, include.only = 'stepAIC')


cancer <- read.csv("../Final Project/FNA_cancer.csv")
attach(cancer)
```





# EDA

## proportions
```{r}
table(cancer$diagnosis)/nrow(cancer)
```


## data melt

```{r}
library(reshape2)

?melt

cancer[-1]

cancer_melt <- melt(cancer[-1])


head(cancer_melt)
```
### All boxplots - no good
```{r}
cancer_melt %>% ggplot(aes(x=variable, y=value))+geom_boxplot(aes(fill=diagnosis))
```

### all variables facet wrapped - also no good
```{r}
cancer_melt %>% ggplot(aes(x=variable, y=value, color=diagnosis)) + geom_boxplot() + 
  facet_wrap( ~ variable, scales = "free") + 
  scale_color_manual(values = c("#4EA2D0","#1F5673"))
```
### plot function definition

```{r}
#"clean boxplot" (no jitter)

#function takes 4 arguemnets (3 variables and one for the plot title, all strings)
facet_plot_function_plain <- function(variable_1,variable_2,variable_3, plot_variable){
  #second half of title string variable
  title_string_2 <- "Values Split by Diagnosis"
  #filters melted dataset to just the requested variables
  cancer_melt %>% 
  filter((variable == variable_1) | (variable == variable_2) | (variable== variable_3)) %>%
  #plots on boxplots in facet wrap fashion.
  ggplot(aes(x=variable, y=value, color=diagnosis)) + geom_boxplot() + 
  facet_wrap( ~ variable, scales = "free") + 
  #custom colors for outline of boxplots and changing the labels of the legend
  scale_color_manual(values = c("#4EA2D0","#1F5673"), name = "Diagnosis", labels=c("Benign", "Malignant")) +
  #labels for title and axis
  labs(title = paste(plot_variable, title_string_2), x="", y="Measurements")
}

#jitter boxplot facet wrap plots

#function takes 4 arguments (3 variables and one for the plot title, all strings)
facet_plot_function_jitter <- function(variable_1,variable_2,variable_3, plot_variable){
  #second half of title string variable
  title_string_2 <- "Values Split by Diagnosis"
  #filters melted dataset to just the requested variables
  cancer_melt %>% 
  filter((variable == variable_1) | (variable == variable_2) | (variable== variable_3)) %>%
  #plots on boxplots in facet wrap fashion.
  ggplot(aes(x=variable, y=value, color=diagnosis)) + geom_boxplot() + 
  #plotting points and jittering them over the boxplots
  geom_point(alpha=0.4, aes(y=value, color=diagnosis), position = position_jitterdodge(jitter.width = .223)) +
  facet_wrap( ~ variable, scales = "free") + 
  #custom colors for outline of boxplots and changing the labels of the legend
  scale_color_manual(values = c("#4EA2D0","#1F5673"), name = "Diagnosis", labels=c("Benign", "Malignant")) +
  #labels for title and axis
  labs(title = paste(plot_variable, title_string_2), x="", y="Measurements")
}
```



## Radius
```{r}
# Mean, SE, and Max are all higher for Malignant
# Mean
cancer %>% ggplot(
  aes(x=radius_mean, y=diagnosis)
) + geom_boxplot() + geom_jitter(alpha=0.5, width = .2)

# Standard Error
cancer %>% ggplot(
  aes(x=radius_se, y=factor(diagnosis, labels = c("Malignant", "Benign")), color=diagnosis)) + 
  geom_boxplot(size=.75, outlier.shape = 1, outlier.color = "black", outlier.size = 3) +
  geom_jitter(alpha=0.4, width = .1, height = .25) +
  scale_color_manual(values = c("#4EA2D0","#1F5673")) +
  labs(title = "Radius", subtitle = "Standard Error",x="Standard Error", y="Diagnosis") + 
  guides(color="none") + theme_minimal()

# Worst
cancer %>% ggplot(
  aes(x=radius_worst, y=diagnosis)
) + geom_boxplot() +
  labs(title = "Title test", subtitle = "subtitle Test",x="radius Worst", y="Diagnosis")
```

###radius melt facet

```{r}
facet_plot_function_plain("radius_mean","radius_se","radius_worst", "Radius")

facet_plot_function_jitter("radius_mean","radius_se","radius_worst", "Radius")

```




## Texture
```{r}
# Mean and Max are slightly higher for Malignant, SE is about the same.
# Mean
cancer %>% ggplot(
  aes(x=texture_mean, y=diagnosis)
) + geom_boxplot()

# Standard Error
cancer %>% ggplot(
  aes(x=texture_se, y=diagnosis)
) + geom_boxplot()

# Worst
cancer %>% ggplot(
  aes(x=texture_worst, y=diagnosis)
) + geom_boxplot()

```
### Texture melt facet

```{r}

facet_plot_function_plain("texture_mean","texture_se" ,"texture_worst", "Texture")

facet_plot_function_jitter("texture_mean","texture_se" ,"texture_worst", "Texture")
```


## Perimeter
```{r}
# Mean, SE, and Max are all higher for Malignant
# Mean
cancer %>% ggplot(
  aes(x=perimeter_mean, y=diagnosis)
) + geom_boxplot()

# Standard Error
cancer %>% ggplot(
  aes(x=perimeter_se, y=diagnosis)
) + geom_boxplot()

# Worst
cancer %>% ggplot(
  aes(x=perimeter_worst, y=diagnosis)
) + geom_boxplot()
```

### perimeter melt facet

```{r}
facet_plot_function_plain("perimeter_mean" ,"perimeter_se", "perimeter_worst", "Perimeter")

facet_plot_function_jitter("perimeter_mean" ,"perimeter_se", "perimeter_worst", "Perimeter")
```




## Area
```{r}

# Mean, SE, and Max are all higher for Malignant
# Mean
cancer %>% ggplot(
  aes(x=area_mean, y=diagnosis)
) + geom_boxplot()

# Standard Error
cancer %>% ggplot(
  aes(x=area_se, y=diagnosis)
) + geom_boxplot()

# Worst
cancer %>% ggplot(
  aes(x=area_worst, y=diagnosis)
) + geom_boxplot()

```
### Area melt facet

```{r}
facet_plot_function_plain("area_mean","area_se","area_worst", "Area")

facet_plot_function_jitter("area_mean","area_se","area_worst", "Area")
```



## Smoothness
```{r}


# Mean and Max are slightly higher for Malignant. SE is about the same.
# Mean
cancer %>% ggplot(
  aes(x=smoothness_mean, y=diagnosis)
) + geom_boxplot()

# Standard Error
cancer %>% ggplot(
  aes(x=smoothness_se, y=diagnosis)
) + geom_boxplot()

# Worst
cancer %>% ggplot(
  aes(x=smoothness_worst, y=diagnosis)
) + geom_boxplot()

```

### smoothness melt facet

```{r}
facet_plot_function_plain("smoothness_mean","smoothness_se","smoothness_worst","Smoothness")

facet_plot_function_jitter("smoothness_mean","smoothness_se","smoothness_worst","Smoothness")
```


## Compactness
```{r}
# Mean, SE, and Max are all slightly higher for Malignant.
# Mean
cancer %>% ggplot(
  aes(x=compactness_mean, y=diagnosis)
) + geom_boxplot()

# Standard Error
cancer %>% ggplot(
  aes(x=compactness_se, y=diagnosis)
) + geom_boxplot()

# Worst
cancer %>% ggplot(
  aes(x=compactness_worst, y=diagnosis)
) + geom_boxplot()

```
### Compactness melt facet

```{r}
facet_plot_function_plain("compactness_mean","compactness_se","compactness_worst", "Compactness")

facet_plot_function_jitter("compactness_mean","compactness_se","compactness_worst", "Compactness")
```


## Concavity
```{r}
# Mean and Max are higher for Malignant. SE is slightly higher.
# Mean
cancer %>% ggplot(
  aes(x=concavity_mean, y=diagnosis)
) + geom_boxplot()

# Standard Error
cancer %>% ggplot(
  aes(x=concavity_se, y=diagnosis)
) + geom_boxplot()

# Worst
cancer %>% ggplot(
  aes(x=concavity_worst, y=diagnosis)
) + geom_boxplot()
```
### concavity melt facet
```{r}
colnames(cancer)
```
"concavity_mean","concavity_se","concavity_worst", "Concavity"
```{r}
facet_plot_function_plain("concavity_mean","concavity_se","concavity_worst", "Concavity")

facet_plot_function_jitter("concavity_mean","concavity_se","concavity_worst", "Concavity")
```


## Concave Points
```{r}
# Mean, SE, and Max are all greater for Malignant
# Mean
cancer %>% ggplot(
  aes(x=concave.points_mean, y=diagnosis)
) + geom_boxplot()

# Standard Error
cancer %>% ggplot(
  aes(x=concave.points_se, y=diagnosis)
) + geom_boxplot()

# Worst
cancer %>% ggplot(
  aes(x=concave.points_worst, y=diagnosis)
) + geom_boxplot()

```

### Concave Points melt facet

```{r}
facet_plot_function_plain("concave.points_mean","concave.points_se","concave.points_worst","Concave Points")

facet_plot_function_jitter("concave.points_mean","concave.points_se","concave.points_worst","Concave Points")
```


## Symmetry 
```{r}

# Mean and Max are slightly higher for Malignant. SE is about the same.
# Mean
cancer %>% ggplot(
  aes(x=symmetry_mean, y=diagnosis)
) + geom_boxplot()

# Standard Error
cancer %>% ggplot(
  aes(x=symmetry_se, y=diagnosis)
) + geom_boxplot()

# Worst
cancer %>% ggplot(
  aes(x=symmetry_worst, y=diagnosis)
) + geom_boxplot()

```
### Symmetry Melt Facet

```{r}
facet_plot_function_plain("symmetry_mean","symmetry_se","symmetry_worst", "Symmetry")

facet_plot_function_jitter("symmetry_mean","symmetry_se","symmetry_worst", "Symmetry")
```


## Fractal Dimension
```{r}

# Mean and SE are pretty similar and Max is slightly higher for Malignant.
# Mean
cancer %>% ggplot(
  aes(x=fractal_dimension_mean, y=diagnosis)
) + geom_boxplot()

# Standard Error
cancer %>% ggplot(
  aes(x=fractal_dimension_se, y=diagnosis)
) + geom_boxplot()

# Worst
cancer %>% ggplot(
  aes(x=fractal_dimension_worst, y=diagnosis)
) + geom_boxplot()


```
### Fractal Dimension melt

```{r}
facet_plot_function_plain("fractal_dimension_mean","fractal_dimension_se","fractal_dimension_worst","Fractal Dimension")

facet_plot_function_jitter("fractal_dimension_mean","fractal_dimension_se","fractal_dimension_worst","Fractal Dimension")
```


# Data Preparation
## Check for NAs
```{r}
colSums(is.na(cancer))
# All columns of X are NA, drop it
# We don't need the ID column, drop it as well
cancer.prep <- cancer %>% select(-c(X, id))
```

## Normalize - needs to be done for KNN classification
```{r}

minmax <- function(x) {
  return ((x - min(x)) / (max(x) - min(x)))
}

cancer.prep <- cancer.prep %>%
  mutate(across(where(is.numeric), minmax))

```




## Set Response as 1, 0 variables - Malign as 1
```{r}
cancer.prep <- cancer.prep %>% 
  mutate(diagnosis = as.factor(ifelse(diagnosis == 'M', 1, 0)))

summary(cancer.prep)

```


```{r}
#head(cancer.prep)

cancer_prep_melt <- melt(cancer.prep)


head(cancer_prep_melt)
```

```{r}
cancer_prep_melt %>% 
  ggplot(aes(x=variable, y=value)) + 
  geom_boxplot(aes(fill=diagnosis)) + 
  theme_bw() + 
  theme(axis.text.x = element_text(angle=45, hjust=1)) +
  labs(title = "Boxplot of Cancer Data",x = "Variable", y = "Value")

```



## Split training and testing sets
```{r}

set.seed(1842)
sets <- cancer.prep %>%
  initial_split(prop = 0.8)

train_set <- sets %>% training()
glimpse(train_set)

test_set <- sets %>% testing()
glimpse(test_set)

form <- as.formula(diagnosis ~ .)

```
### malignant/benign proportions

### training set
```{r}
#Set Response as 1, 0 variables - Malign as 1
table(train_set$diagnosis)/nrow(train_set)
```


### test set

```{r}
table(test_set$diagnosis)/nrow(test_set)
```

# Models

## Decision Tree
```{r}

# 92.1% Accuracy, 96.8% Sensitivity
# Fit the tree
dt <- rpart(form, data = train_set, cp=0, minsplit=2)

# Plot the tree
plot(as.party(dt))

# Check CP options
plotcp(dt)

# Prune
dt.pruned <- prune(dt, cp=0.019)

# Plot the pruned tree
plot(as.party(dt.pruned))

# Predict on the test set
dt.preds <- predict(dt.pruned, test_set, type='class')

# Confusion matrix and performance metrics
confusionMatrix(dt.preds, as.factor(test_set$diagnosis), positive = '1')

```




## Random Forest
```{r}
# 94.7% Accuracy, 96.9% Sensitivity
# Use variables (mtry) around sqrt of total predictors (33)
rf <- randomForest(form, data=train_set, mtry=6, ntree=1000)
print(rf)

# Check variable importance
varImpPlot(rf, sort=TRUE) 

# Predict on test set
rf.preds <- predict(rf, test_set, type='class')

# Confusion matrix and performance metrics
confusionMatrix(rf.preds, as.factor(test_set$diagnosis), positive = '1')

```

### Hyper-tuning mtry
```{r}
set.seed(1842)
#number of mtry
mtry_values <- seq(1,30, by=1)

#vectors for storing model metrics
accuracy_results_mtry <- rep(0)
sensitivity_results_mtry <- rep(0)
specificity_results_mtry <- rep(0)

#loop for each number of mtry for the model
for(i in seq_along(mtry_values)){
  mtry <- mtry_values[i]
  #fitting model with various values of mtry
  rf_temp <- randomForest(form, data=train_set, mtry=mtry, ntree=1000)
  #making predictions
  rf_temp_preds <- predict(rf_temp, test_set, type='class')
  
  #creating confusion matrix
  test_table <- table(rf_temp_preds, test_set$diagnosis) 
  
  #true negatives
  tn <- test_table[1]
  #flase positives
  fp <- test_table[2]
  #flase negatives
  fn <- test_table[3]
  #true positives
  tp <- test_table[4]
  
  #creating accuracy, sensitivity, specificity metrics
  accuracy_results_mtry[i] <- ((tp+tn)/(tp+fp+tn+fn))
  sensitivity_results_mtry[i] <- (tp)/(tp+fn)
  specificity_results_mtry[i] <- (tn)/(tn+fp)
}

#data frame for accuracy, sensitivity, specificity results
mtry_results <- data.frame(mtry_values, accuracy_results_mtry, sensitivity_results_mtry, specificity_results_mtry)
# finding max index for accuracy, sensitivity, specificity results
mtry_acc_max <- which.max(mtry_results$accuracy_results_mtry)
mtry_sens_max <- which.max(mtry_results$sensitivity_results_mtry)
mtry_spec_max <- which.max(mtry_results$specificity_results_mtry)

#displaying max result for accuracy, sensitivity, specificity
mtry_results[mtry_acc_max,]
mtry_results[mtry_sens_max,]
mtry_results[mtry_spec_max,]

#plot mtry accuracy results
plot(x=mtry_values, y=accuracy_results_mtry, type='l', col='red',xlab = "mtry", ylab = "results", xlim = c(0, 35), ylim = c(.9, 1))
#plot mtry sensitivity results
lines(mtry_values, sensitivity_results_mtry, col='blue', type='l')
#plot mtry specificity results
lines(mtry_values, specificity_results_mtry, col='green', type='l')
#legend and title
legend("bottomright", legend = c("Accuracy", "Sensitivity", "Specificity"), col = c("red","blue","green"), lty=1)
title(main='values of mtry')

```

### Hyper-tuning number of trees
```{r}
set.seed(1842)
#number of trees from 50 to 2000
trees_values <- seq(50,2000, by=50)

#vectors for storing model metrics
accuracy_results_trees <- rep(0)
sensitivity_results_trees <- rep(0)
specificity_results_trees <- rep(0)

#loop for each number of trees for the model
for(i in seq_along(trees_values)){
  #
  trees <- trees_values[i]
  #fitting random forest model with mtry of 15 and various number of trees
  rf_temp <- randomForest(form, data=train_set, mtry=15, ntree=trees)
  #making predictions
  rf_temp_preds <- predict(rf_temp, test_set, type='class')
  
  #creating confusion matrix
  test_table <- table(rf_temp_preds, test_set$diagnosis)
  
  #true negatives
  tn <- test_table[1]
  #flase positives
  fp <- test_table[2]
  #flase negatives
  fn <- test_table[3]
  #true positives
  tp <- test_table[4]
  
  #creating accuracy, sensitivity, specificity metrics
  accuracy_results_trees[i] <- ((tp+tn)/(tp+fp+tn+fn))
  sensitivity_results_trees[i] <- (tp)/(tp+fn)
  specificity_results_trees[i] <- (tn)/(tn+fp)
}

#data frame for accuracy, sensitivity, specificity results
trees_results <- data.frame(trees_values, accuracy_results_trees, sensitivity_results_trees, specificity_results_trees)
# finding max index for accuracy, sensitivity, specificity results
trees_acc_max <- which.max(trees_results$accuracy_results_trees)
trees_sens_max <- which.max(trees_results$sensitivity_results_trees)
trees_spec_max <- which.max(trees_results$specificity_results_trees)

#displaying max result for accuracy, sensitivity, specificity
trees_results[trees_acc_max,]
trees_results[trees_sens_max,]
trees_results[trees_spec_max,]

#plot ntree accuracy results
plot(x=trees_values, y=accuracy_results_trees, type='l', col='red',xlab = "trees", ylab = "results", xlim = c(0, 2050), ylim = c(.9, 1))
#plot ntree sensitivity results
lines(trees_values, sensitivity_results_trees, col='blue', type='l')
#plot ntree specificity results
lines(trees_values, specificity_results_trees, col='green', type='l')
#legend and title
legend("bottomright", legend = c("Accuracy", "Sensitivity", "Specificity"), col = c("red","blue","green"), lty=1)
title(main="Number of trees by 50")
abline(v=450, lty=2)

```

### Hyper-tuned random forest model
```{r}

set.seed(1842)
RandomForest_tuned <- randomForest(form, data=train_set, mtry=15, ntree=450)
rf_tuned_preds <- predict(rf_tuned, test_set, type='class')

rf_tuned_table <- table(rf_tuned_preds, test_set$diagnosis)
confusionMatrix(rf_tuned_table, positive = '1')

```

```{r}
varImpPlot(RandomForest_tuned)
```





## KNN 
```{r}
# 95.6% Accuracy, 90.6% Sensitivity
# Use k around sqrt of n (569)
knn.24 <- knn(train=train_set[-1], test=test_set[-1], cl=train_set$diagnosis, k=24)

# Confusion matrix and performance metrics
confusionMatrix(knn.24, as.factor(test_set$diagnosis), positive = '1')

```

### Hyper-tuning k
```{r}
set.seed(1842)
#vector for k values from 1 to 101
k_values <- seq(1,101, by=2)

#vectors for storing model metrics
accuracy_results_knn <- rep(0)
sensitivity_results_knn <- rep(0)
specificity_results_knn <- rep(0)

#loop for each value of k for the model
for(i in seq_along(k_values)){
  #establishing the loops value of k
  k <- k_values[i]
  #fitting model with k
  knn_temp <- knn(train=train_set[-1], test=test_set[-1], cl=train_set$diagnosis, k=k)
  
  #creating confusion matrix
  test_table <- table(knn_temp, test_set$diagnosis)
  
  #true negatives
  tn <- test_table[1]
  #flase positives
  fp <- test_table[2]
  #flase negatives
  fn <- test_table[3]
  #true positives
  tp <- test_table[4]
  
  #creating accuracy, sensitivity, specificity metrics
  accuracy_results_knn[i] <- ((tp+tn)/(tp+fp+tn+fn))
  sensitivity_results_knn[i] <- (tp)/(tp+fn)
  specificity_results_knn[i] <- (tn)/(tn+fp)
  
}

#data frame for accuracy, sensitivity, specificity results
knn_results <- data.frame(k_values, accuracy_results_knn, sensitivity_results_knn, specificity_results_knn)
# finding max index for accuracy, sensitivity, specificity results
knn_acc_max <- which.max(knn_results$accuracy_results)
knn_sens_max <- which.max(knn_results$sensitivity_results)
knn_spec_max <- which.max(knn_results$specificity_results_knn)

#displaying max result for accuracy, sensitivity, specificity
knn_results[knn_acc_max,]
knn_results[knn_sens_max,]
knn_results[knn_spec_max,]

#plot accuracy results
plot(x=k_values, y=accuracy_results_knn, type='l', col='red',xlab = "K", ylab = "results", xlim = c(0, 103), ylim = c(.8, 1))
#plot sensitivity results
lines(k_values, sensitivity_results_knn, col='blue', type='l')
#plot specificity results
lines(k_values, specificity_results_knn, col='green', type='l')
#legend and title
legend("right", legend = c("Accuracy", "Sensitivity", "Specificity"), col = c("red","blue", "green"), lty=1)
title(main="Various Values of K")
abline(v=3, lty=2)

```

### Hyper-tuned knn model
```{r}

set.seed(1842)
#fitting model with k=3
knn_3 <- knn(train=train_set[-1], test=test_set[-1], cl=train_set$diagnosis, k=3)
#k=3 model results
confusionMatrix(knn_3, as.factor(test_set$diagnosis), positive = '1')

```

##Logistic Regression
```{r}

# 93% Accuracy, 90.6% Sensitivity
# Create the full model
# Does not converge even after stepAIC
lr.full <- glm(form, data=train_set, family=binomial, control = list(maxit = 100))

stepAIC(lr.full)

lr.reduced <- glm(formula = diagnosis ~ radius_mean + texture_mean + smoothness_mean + 
      compactness_mean + concavity_mean + fractal_dimension_mean + 
      area_se + compactness_se + symmetry_se + fractal_dimension_se + 
      radius_worst + concave.points_worst + symmetry_worst + fractal_dimension_worst, 
    family = binomial, data = train_set, control = list(maxit = 50))

# Predict on test set
lr.preds <- as.factor(round(predict(lr.reduced, test_set, type='response')))

# Confusion matrix and performance metrics
confusionMatrix(lr.preds, as.factor(test_set$diagnosis), positive = '1')



```

